
%% dedicatoria
\begin{dedicatory}
    Dedicated to whoever is reading this document.
\end{dedicatory}

%% agradecimientos
\begin{acknowledgment}
This is probably the hardest section to write. Honestly. I'll begin stating that people here will be named in no particular order, as they are mentioned as soon as they pop into my head. There is a lot of people that I want to thank. To my parents Eraclio, Licean and my little sister Licean ---yes, there is another---, for being there even if I have been a terrible son from the beginning. \\

To all the staff of my university's  CS department ---both old and new people--- with a special mention to Marcela Pacheco and Ruth Garrido, which have been supporting me from the start. \\

I would also like to acknowledge Felipe Astroza from which I started looking at him as a rival and now we're even working together in the same place. I've always admired how he done things and he did not only taught me probably most of the quirks that I have even today when it comes to programming, but also that knowledge should be for everyone.\\

A shoutout to Brayan Gonzalez, my best friend which I met during first years of studying CS for all his support during this time and for always being there, even for the silliest of things. In these times of quarantine I must acknowledge that I miss you a lot, specially those drunken programming moments.\\

Awarded with a unique special mention is Rapa (my advisor). Both of us procrastinate a lot, we leave everything for last moment, we argue and yell each other a lot. Yet here we are. I'm thankful for the transmited experience during all this time and also for being a reliable friend. As I said before, he is the worst advisor that I could choose, yet the only one that it is up to the task.\\

Also I want to thank all the people from the former \#weebs IRC channel, now a rather huge group of people for supporting me when I decided to go to live in Santiago and start as a VJ. A special mention to Jorge Solar and Felipe Rojas for being so encouraging when I was starting.\\

A special mention to Daniel Hedencrona for all his support during this time and for being a reliable friend for more than 7 years so far. He is not only one of the few guys which can understand my hobbies, but also knowledge on many topics. I mean, years ago we discussed implementations for a duck hunt AI just for fun --- and because he was learning to implement bayesian trees.\\

Finally I would like to thank Benjamin Haller from the Messer Lab not only for being the source of inspiration for this work, but also for all his disposition during the development of the initial prototypes of bIIQS which are now installed onto SLiM. \\

To my students from the competitive programming workshop, for allowing me being their coach, even if we did not won any award. And also to the local ICPC staff for allowing me being part of such a wonderful event. You are so cool, really.\\

To the trio that teased me for years until I said that I was going to get my degree: Cristian Ruz, Carla Gallardo and Alejandro Sazo. I really miss going to eat peruvian food with you guys. Also I want to thank Martin Gutierrez and Rodolfo Allendes because if it were not for their influence during my time studying bioinformatics I would have given up on studying engineering.\\

Talking about bioinformatics, a big thanks and hugs to my friend Samuel Ortega, which went to US to study and still does not return. Hope he does not get any weird virus over there.\\

I also want to acknowledge to all the people whom I'm no longer in contact anymore. Some of them because of fights, others are no longer with us and many others because I am usually afraid of talking with other people. It's not like I have forgotten about you.\\
\end{acknowledgment}